### 2020-01-16 Manuscript.Rmd
markdown file

---
title: 
author: Tina & Marc
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_notebook:
    fig_caption: yes
    keep_tex: yes
    highlight: espresso
    number_sections: no
  html_document:
    df_print: paged
  word_document:
    reference_docx: Manuscript.docx
urlcolor: blue
fontsize: 10pt
geometry: margin=1in
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{float}
- \usepackage{color}
linestretch: 1
bibliography: /Users/admin/Desktop/nwork/simulation/TM.bib
csl: /Users/admin/Desktop/nwork/simulation/styles-master/styles-master/plos.csl 
---

@@

Most comparative studies of functional genomics have used pairwise comparisons. Yet it has been shown that this can provide biased results, since genes, like species, are phylogenetically related. Phylogenetic comparative methods should allow to correct for this, but they depend on strong assumptions, including unbiased tree estimates relative to the hypothesis being tested. Such methods have recently been used to test the "Ortholog Conjecture", the hypothesis that functional evolution is faster in paralogs than orthologs. Whereas pairwise comparisons of tissue specificity index ($\tau$) provided support for the ortholog conjecture, it was not supported using phylogenetic independent contrasts. We find that the gene trees used suffer from important biases, due to the inclusion of trees with no duplication nodes, to the relative age of speciations and duplications, to systematic differences in branch lengths, and to non-Brownian motion of tissue-specificity on many trees. We find some support for the ortholog conjecture, but especially that incorrect implementation of phylogenetic method in empirical gene trees with duplications can be problematic.

@@

There is a lot of interest in understanding the evolution of gene function. Comparing functional genomics results offers a way to do this. In most cases, it is done by comparing pairs of genes, and notably pairs of orthologs (homologs derived by speciation) and pairs of paralogs (homologs derived by duplication). A drawback of such pairwise comparisons is that they neglect the evolutionary history of the genes, and thus violate the statistical presumption of independence of observations. Phylogenetic comparative methods have been suggested to correct for this. Recent results indicate that whereas pairwise comparisons support stronger functional divergence of paralogs than of orthologs, phylogenetic analyses would not. Re-analyses of these data show that gene trees are biased in such a way as to cause biases in phylogenetic methods when there are gene duplication. This can led to erroneous results, and have serious consequences in biological data interpretation.

@@

The "Ortholog Conjecture", a cornerstone of phylogenomics, has become a topic of debate in recent years [@RN771;@RN774;@RN790;@RN793;@RN794;@RN762;@RN760;@RN770;@RN792]. Dealing with the roles of gene duplications in functional evolution, the ortholog conjecture is routinely used by both experimental and computational biologists in predicting, or understanding gene function. According to this model, orthologs (i.e. homologous genes which diverged by a speciation event) retain equivalent or very similar functions, whereas paralogs (i.e. homologous genes which diverged by a duplication event) share less similar functions [@RN771]. This is linked to the hypothesis that paralogs evolve more rapidly. This hypothesis was challenged by results suggesting that paralogs would be functionally more similar than orthologs [@RN774]. Such findings not only raised questions on the evolutionary role of gene duplication but also questioned the reliability of using orthologs to annotate unknown gene functions in different species [@RN844]. Several more recent studies [@RN770;@RN790;@RN793;@RN794;@RN762] found support for ortholog conjecture, mostly based on comparisons of gene expression data.

While all previous studies of the ortholog conjecture had used pairwise comparisons of orthologs and paralogs, a recent article suggested that this was flawed, and that phylogenetic comparative methods should be used [@RN760]. In the field of comparative biology, most of the studies still rely on the pairwise method including the testing of the ortholog conjecture. Ignorance of phylogenetic structure has been reported to underestimate the fundamental assumption of independent observations in statistics [@RN773;@RN807]. A solution is to use phylogeny-based methods to investigate such evolutionary patterns [@RN782;@RN807;@RN779;@RN780;@RN773;@RN850;@RN849]. There are three main such phylogenetic methods: Phylogenetic Independent Contrast (PIC) [@RN773], Phylogenetic Generalized Least-Square (PGLS) [@RN850], and Monte Carlo computer simulation [@RN849]. PIC is widely adopted for its relative simplicity, and its applicability to a wide range of statistical procedures [@RN781;@RN760]. The performance of PIC relies on three basic assumptions: a correct tree topology; accurate branch lengths; and trait evolution following a Brownian model (where trait variance accrues as a linear function of time)  [@RN781;@RN782;@RN783;@RN807;@RN780;@RN779;@RN816]. If any of these assumptions is incorrect, this can lead to incorrect interpretation of results. This is probably why the application of such phylogenetic methods is still limited, and debated even after being introduced about three decades ago [@RN781;@RN782;@RN783;@RN807;@RN780]. Dunn et al. [@RN760] took an innovative approach in applying PIC to compare the divergence rates between two different events ("speciation" and "duplication") to test the ortholog conjecture. However, such an application might be problematic since the time of occurrence of gene duplication, one of the two types of events compared, is unknowable by external information (e.g. no fossil evidence). Therefore, further study is required to understand why Dunn et al. [@RN760] obtained results which are inconsistent with most studies by using the phylogenetic method. It is possible that all the conclusions drawn by previous studies on gene duplication are incorrect due to overlooking phylogenetic tree structure. If so, it should be well supported.

We re-examined the data of Dunn et al., after reproducing it using the resources and scripts provided by the authors [@RN760]. Our reanalysis highlights potential problems associated with phylogenetic independent contrasts when applied to the impact of gene duplication. Finally, with proper controls, the phylogenetic method supports the ortholog conjecture.

@@

@@

__Issues with naive use of Phylogenetic Independent Contrasts (PICs)__

@@

To understand their results, we first reanalyzed the data of Dunn et al. [@RN760]. We were able to reproduce all the results published in their article by running the code, which they clearly supplied. Dunn et al. reported a non-significant result for the PIC under the null simulation, using a Wilcoxon one-tailed rank test to check if the contrasts of duplication events are higher than the contrasts of speciation events (*P* = $`r format(wilcox_one_tailed1, digits= 3, scientific = FALSE)`$). Surprisingly, the PIC rejects the null hypothesis on the null simulations with a Wilcoxon two-tailed rank test (Fig 1A), with significant support for higher contrasts after speciation than duplication. This was robust to repeating the simulations with different random seed number (S1 Fig). This indicates that neither of the methods, PIC or pairwise, worked properly for these calibrated trees, since both rejects the null when simulations are performed under null.

@@

The purpose of applying a phylogenetic method is to generate independent statistical data points for traits, which are otherwise not independent enough because of their phylogenetic relatedness. Such statistical non-independence among species trait values can be measured by "phylogenetic signal" [@RN819; @RN768; @RN800; @RN820;@RN826]. Blomberg's K and Pagel's $\lambda$ are widely used methods to capture such phylogenetic signal [@RN819;@RN768]. Dunn et al. [@RN760] used Blomberg's K in their study. In case of Blomberg's K the value ranges from 0 to $\infty$, where a value of 0 indicates no phylogenetic signal, a value of 1 indicates trait evolution according to a Brownian model (BM), values between 0 and 1 indicate that distant relatives are more similar than expected under BM, and values larger than 1 indicate that close relatives are more similar than expected under BM [@RN819; @RN768; @RN800; @RN820;@RN826]. With a cutoff of K > 0.551, Dunn et al. [@RN760] obtained only `r subset.trees` out of `r length(gene_trees_calibrated)` calibrated trees with strong phylogenetic signal. Using a cut-off of *P* < 0.05 together with K > 0.551 leads to 1135 trees, which produce similar results (S3 Fig) to that set of `r subset.trees` trees. We continued analyses with the `r subset.trees` trees of Dunn et al. [@RN760] for consistency. In any case, it is notable that trait ($\tau$) values are independent of phylogeny for the majority of the gene trees. For these `r (length(gene_trees_calibrated)-subset.trees)` trees, implementation of PIC to produce statistically independent data points might not be necessary. Moreover, it can be misleading if the contrasts are not checked for adequate standardization as per BM after applying PIC (discussed later). The phylogenetic method still rejects the null hypothesis under null simulations for those `r subset.trees` trees (Fig 1B), showing that the problem is not simply due to low phylogenetic signal.

@@

```{r Detailed_analyses_on_empirical_data, echo=F, message=F, warning=F}

@@

Dunn et al. [@RN760] used `r nrow(calibration_times_focal)` speciation time points to calibrate gene trees. The oldest speciation calibration was at `r max(calibration_times_focal$age)` My.  All other calibrated nodes were duplication nodes leading to obtain ~`r length(duplication.empirical)` unique duplication time points for the `r length(gene_trees_calibrated)` calibrated trees [@RN760]. Out of these, `r length(duplication.empirical[duplication.empirical > max(speciation.empirical)])` time points preceded the oldest speciation node age. Surprisingly, the calibrated node age of the oldest duplication event was found to be `r as.integer(max(duplication.empirical))` My, that is, 2600 times older than the Earth. This is indicative of the difficulty of estimating the age of ancient duplications by phylogenetic methods. Those `r length(duplication.empirical[duplication.empirical > max(speciation.empirical)])` high duplication node ages eventually led to much larger expected variances for gene duplication events (median expected variance for duplication events preceding the oldest speciation events: `r round(median(Exp_var.duplication.precede.speciation),0)`, median expected variance for speciation events: `r round(median(Exp_var.speciation),0)`). This explains why the mean paralog distance was ~`r round((paralog_distance_mean/ortholog_distance_mean),1)` times higher than that of the mean ortholog distance (S4A Fig). This data distribution also makes it appear as if no speciation happened before `r max(speciation.empirical)` My, a problem shared with the pairwise analysis of the same data [@RN762]. There are also branch length issues for duplication events younger than the oldest speciation events, including very short branches and negative branch lengths. All this led to obtain abnormally high duplication contrasts for the empirical data. To limit such problems, Dunn et al. [@RN760] removed node contrasts higher than 0.5 on the empirical data (this does not impact Fig 1, as the simulation data never has such high contrasts). Following this practice, there are still higher expected variances following gene duplication events (S4A and S4B Figs). In the null simulations only the $\tau$ values are simulated, while the branch lengths (hence the variances) are taken from the empirical data, and thus share its biases. This explains why contrasts are lower for duplications than for speciations under null simulations as well as with empirical data.

@@

  ## All calibrated trees with permuted actual Tau of the tips 
  Tau_randomized_tree <- mclapply(gene_trees_pic,shuffling_tau,mc.cores = cores)

@@

  ## All calibrated trees with permuted actual internal node events
  Event_randomized_pic_tree <- mclapply(gene_trees_pic, shuffling_event, mc.cores = cores)

@@

  ## Histogram to show distribution of proportion of events for all calibrated trees
  event_histogram <- ggplot2_histogram_mod(data=plot_data, xName='proportion_event',
    groupName='Event',legendPosition="right",
    alpha=1, addDensity=TRUE,
    addMedianLine=TRUE, medianLineColor=c("#F8766D","#00BFC4"), medianLineSize=1) +
    theme_classic() +
    ggtitle(expression(bold("D. Event proportion plot")))+
    theme(plot.title = element_text(face="bold"))+
    xlab("Proportion of events") + 
    ylab("Density") +
    theme(legend.title=element_blank(),legend.position=c(0.9,0.8),legend.text=element_text(size=8)) +
    theme(axis.text=element_text(size=10,face="bold")) +
    theme(plot.title = element_text (face="bold", 
                              colour="black", lineheight=1.0),
          axis.title.x=element_text( face="bold",
                              colour="black", hjust=0.5),
          axis.title.y=element_text(face="bold",
                              colour="black", vjust=0.5, angle=90))

@@

We used randomization tests to assess whether the results of different analyses of the empirical dataset are reliable and unbiased. In a first randomization test, we randomized the trait values (i.e. randomly permuting the $\tau$ values) across the tips of each tree without altering the node events of the trees. We then computed the contrasts for the speciation and duplication events of the trees when there is no relation between trait values and phylogenetic relationships. When we compared the nodes contrasts of speciation and duplication events of these randomized trees (Fig 2A), we found the same pattern as reported for the empirical gene trees by Dunn et al. [@RN760]. This shows that the PIC applied to these data is not measuring phenotypic evolution. It confirms that results are driven by their large differences in branch lengths (i.e. in expected variances) (Fig 2B), as on simulated null data. Any effect of trait divergence rates of speciation and duplication events is masked by this branch length difference. This violates the basic assumption of applicability of the PIC method to Brownian trait evolution. To remove the problem of difference in expected variances of the two events, we performed a second randomization test: we kept the original $\tau$ value for tips but randomly shuffled the events (duplication or speciation) of internal nodes of the empirical gene trees to maintain the original proportions of speciation and duplication events. The resulting trend (Fig 2C) still resembled the empirical gene trees data of Dunn et al. [@RN760]. This appears due to the fact that the majority of the tree events are speciation events (Fig 2D) with speciation node ages $\leq$ 296 My. Most of the trees with many duplication events on the other hand have ancient duplication events (duplication node age > 296 My) for which the evolutionary rates of duplication are often masked by the effect of longer branch lengths. Hence, opposite to our expectation, the calibrated trees with no or few duplications have higher overall nodes contrast (apparent fast evolution) than trees with many duplications (apparent slow evolution). This might be due to greater difficulty in detecting paralogs for fast evolving genes. Therefore, reshuffling of the events may not change the observed pattern of higher speciation contrasts than duplication contrasts.

@@

```{r Fig2, dpi=400,fig.width=8, fig.height=8, fig.cap="**Fig 2**: Analyses on calibrated empirical gene trees of Dunn et al. [@RN760]. *P* values are from  Wilcoxon two-tailed tests. (A) Randomly shuffling the $\\tau$ values of the tips for 8520 gene trees does not alter the trend of empirical result [@RN760] showing an opposite trend to the ortholog conjecture. (B) The figure indicates the expected variance is much higher for duplication than speciation events irrespective of the number of tips considered for the study. (C) Using the original $\\tau$ data, if we permute the events (Speciation/Duplication/NA) of the nodes, the trend of result still resembles the empirical result obtained by the recent study [@RN760]. (D) The proportions of speciation events are much higher than duplication events for all time-calibrated trees. The dotted line represents the median proportion of both events. This plot shows a high frequency of trees have no/few duplication events."}

  ## Plotting the randomization data result for all calibrated trees 
  ## 1st plot
  jitterplot_randomized_tau <- 
    ggplot(nodes_contrast_random,aes(x=Event, y=pic_abs_random, fill=Event)) + 
    geom_jitter(aes(colour = Event),position=position_jitter(0.02), alpha=0.02) +
    geom_text(data = median_nodes_contrast_random, aes(label = pic.round),fontface = 2,size = 4, 
              hjust=0.5,vjust =-2.5)+
    geom_crossbar(data=median_nodes_contrast_random, aes(ymin = pic_abs_random, ymax = pic_abs_random),
                  size=0.2,col= "black", width = .2)+
    xlab( NULL ) +
    ylab(expression(bold("Phylogenetic Independent Contrasts (PICs)"))) +
    ylim(0, 0.4) +
    theme_classic()+
    theme(legend.title=element_blank(),legend.position=c(1.9,1.9)) +
    theme(axis.text=element_text(size=10,face="bold")) +
    theme(legend.text=element_text(size=10,face="bold")) +
    annotate("text", x = 1.5, y = 0.3, label= wilcox_2_tailed.r, fontface = 4)+
    ggtitle(expression(bold(paste("A. Plot using randomized trait, ", tau))))+
    theme(plot.title = element_text(face="bold"))

@@

  jitterplot_randomized_event <- 
    ggplot(nodes_contrast_random_events,aes(event_new, y=pic, fill=event_new)) + 
    geom_jitter(aes(colour = event_new),position=position_jitter(0.02), alpha=0.02) +
    geom_text(data = median_random_event, aes(label = pic.round),fontface = 2,size = 4, 
              hjust=0.5,vjust = -2.5)+
    geom_crossbar(data=median_random_event, aes(ymin = pic, ymax = pic),
                  size=0.2, col= "black",width = .2)+
    xlab( NULL ) +
    ylab( NULL) +
    ylim(0, 0.4) +
    theme_classic()+
    theme(legend.title=element_blank(),legend.position=c(1.9,1.9)) +
    theme(axis.text=element_text(size=10,face="bold")) +
    theme(legend.text=element_text(size=10,face="bold")) +
    annotate("text", x = 1.5, y = 0.3, label= wilcox_2_tailed_revent, fontface = 4)+
    ggtitle(expression(bold(paste("C. Plot using randomized events"))))+
    theme(plot.title = element_text(face="bold"))

@@

__The ortholog conjecture when $\tau$ evolution follows a Brownian model__

@@

>>>>

---
title: 
author: Tina & Marc
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_notebook:
    fig_caption: yes
    keep_tex: yes
    highlight: espresso
    number_sections: no
  html_document:
    df_print: paged
  word_document:
    reference_docx: 
urlcolor: blue
fontsize: 10pt
geometry: margin=1in
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{float}
- \usepackage{color}
linestretch: 1
bibliography: /Users/admin/Desktop/nwork/simulation/TM.bib
csl: /Users/admin/Desktop/nwork/simulation/styles-master/styles-master/plos.csl 
---

@@

Most comparative studies of functional genomics have used pairwise comparisons. Yet it has been shown that this can provide biased results, since genes, like species, are phylogenetically related. Phylogenetic comparative methods should allow to correct for this, but they depend on strong assumptions, including unbiased tree estimates relative to the hypothesis being tested. Such methods have recently been used to test the "ortholog conjecture", the hypothesis that functional evolution is faster in paralogs than in orthologs. Whereas pairwise comparisons of tissue specificity index ($\tau$) provided support for the ortholog conjecture, phylogenetic independent contrasts did not. We find that the gene trees used suffer from important biases, due to the inclusion of trees with no duplication nodes, to the relative age of speciations and duplications, to systematic differences in branch lengths, and to non-Brownian motion of tissue-specificity on many trees. We find some support for the ortholog conjecture, but especially that incorrect implementation of phylogenetic method in empirical gene trees with duplications can be problematic.

@@

There is a lot of interest in understanding the evolution of gene function. Comparing functional genomics results offers a way to do this. In most cases, it is done by comparing pairs of genes, and notably pairs of orthologs (homologs derived by speciation) and pairs of paralogs (homologs derived by duplication). A drawback of such pairwise comparisons is that they neglect the evolutionary history of the genes, and thus violate the statistical presumption of independence of observations. Phylogenetic comparative methods have been suggested to correct for this. Recent results indicate that whereas pairwise comparisons support functional divergence of paralogs than of orthologs, phylogenetic analyses would not. Re-analyses of these data show that gene trees are biased in such a way as to cause biases in phylogenetic methods when there are gene duplication. This can led to erroneous results, and have serious consequences in biological data interpretation.

@@

The "ortholog conjecture", a standard model of phylogenomics, has become a topic of debate in recent years [@RN771;@RN774;@RN790;@RN793;@RN794;@RN762;@RN760;@RN770;@RN792;@RN870]. Dealing with the roles of gene duplications in functional evolution, the ortholog conjecture is routinely used by both experimental and computational biologists in predicting or understanding gene function. According to this model, orthologs (i.e. homologous genes which diverged by a speciation event) retain equivalent or very similar functions, whereas paralogs (i.e. homologous genes which diverged by a duplication event) share less similar functions  [@RN771]. This is linked to the hypothesis that paralogs evolve more rapidly. This hypothesis was challenged by results suggesting that paralogs would be functionally more similar than orthologs [@RN774]. Such findings not only raised questions on the evolutionary role of gene duplication but also questioned the reliability of using orthologs to annotate unknown gene functions in different species [@RN844]. Several more recent studies [@RN790;@RN793;@RN794;@RN762] found support for the ortholog conjecture, mostly based on comparisons of gene expression data.

\vspace{0.05cm}
While all previous studies of the ortholog conjecture had used pairwise comparisons of orthologs and paralogs, a recent article suggested that this was flawed, and that phylogenetic comparative methods should be used [@RN760]. Ignorance of phylogenetic structure has been reported to underestimate the fundamental assumption of independent observations in statistics [@RN807]. A solution is to use phylogeny-based methods to investigate such evolutionary patterns [@RN782;@RN807;@RN779;@RN780;@RN850;@RN849]. There are three main such phylogenetic methods: Phylogenetic Independent Contrast (PIC) [@RN807], Phylogenetic Generalized Least-Square (PGLS) [@RN850], and Monte Carlo computer simulation [@RN849]. PIC is widely adopted for its relative simplicity, and its applicability to a wide range of statistical procedures [@RN781;@RN760]. The performance of PIC relies on three basic assumptions: a correct tree topology; accurate branch lengths; and trait evolution following a Brownian model (where trait variance accrues as a linear function of time) [@RN781;@RN782;@RN783;@RN807;@RN780;@RN779;@RN816]. If any of these assumptions is incorrect, this can lead to incorrect interpretation of results. Dunn et al. [@RN760] took an innovative approach in applying PIC to compare the divergence rates between two different events ("speciation" and "duplication") to test the ortholog conjecture, and performed extensive analyses in support of their results. However, such an application might be problematic since the time of occurrence of gene duplication, one of the two types of events compared, is unknowable by external information (e.g. no fossil evidence). Therefore, further study is required to understand why Dunn et al. [@RN760] obtained results, which are inconsistent with previous studies. It is possible that all the conclusions drawn by previous studies on gene duplication are incorrect due to overlooking phylogenetic tree structure. If so, it should be well supported.

\vspace{0.05cm}
We re-examined the data of Dunn et al., after reproducing their results using the resources and scripts provided by the authors [@RN760]. Our reanalysis highlights potential problems associated with phylogenetic independent contrasts when applied to the impact of gene duplication. Finally, with proper controls, the phylogenetic method supports the ortholog conjecture on the data used by Dunn et al.

@@

	## Retrieving speciation and duplication PICs for empirical data of Dunn et al.
  speciation_empirical <- nodes_contrast$pic[which(nodes_contrast$Event=="Speciation" & nodes_contrast$pic < 0.5)]
  duplication_empirical <- nodes_contrast$pic[which(nodes_contrast$Event=="Duplication" & nodes_contrast$pic < 0.5)] ## for them

   ## Performing two sided Wilcoxon test on empirical data 
  wilcox_two_tailed_empirical <- two_tailed_wilcox(speciation_empirical,duplication_empirical)

@@

__Issues with naive application of Phylogenetic Independent Contrasts (PICs)__

@@

Many studies of functional evolution following gene duplication have used a pairwise approach [@RN762;@RN771;@RN790;@RN793;@RN794;@RN774]. All such pairwise studies, except two [@RN774; @RN870], have supported the ortholog conjecture [@RN762;@RN771;@RN790;@RN793;@RN794]. In this context, Dunn et al. [@RN760] have made a relevant argument that the test should be done in a phylogenetic framework, since closely related species or genes tend to share more similar traits. Parsing empirical gene trees from ENSEMBL Compara v.75, using the tissue-specificity data ($\tau$) from [@RN762] as the phenotypic trait for 8 vertebrate species, and applying PIC, they reported evidence in contradiction with the ortholog conjecture [@RN760]. To understand the incongruence between results of PIC and of pairwise methods, they performed simulations of $\tau$ on their `r length(gene_trees_calibrated)` time calibrated trees under the OC (ortholog conjecture), and under a null of uniform Brownian motion. Under the null simulation, the same trait evolutionary rate is used for speciation and duplication events ($\sigma$^2^~$\tau$-duplication~ = $\sigma$^2^~$\tau$-speciation~), thus the phylogenetic contrasts or pairwise correlations of different events should be drawn from the same distribution. For the OC simulation, a higher rate of trait evolution is applied following gene duplication relative to speciation; they used $\sigma$^2^~$\tau$-duplication~ = 2 \*  $\sigma$^2^~$\tau$-speciation~. Under OC, for PIC we expect higher duplication contrasts than speciation contrasts, and for pairwise comparisons we expect a higher correlation coefficient of ortholog pairs than of paralog pairs. The simulation results of Dunn et al. [@RN760] indicated that the pairwise method could not distinguish the two scenarios (null and OC), while the PIC method was able to recover expected trends. As the PIC on empirical data resembled their null simulation, they questioned both the use of pairwise methods, and the support for the ortholog conjecture from tissue specificity data.

To understand their results, we first reanalyzed the data of Dunn et al. [@RN760]. We were able to reproduce all the results published in their article by running the code, which they clearly supplied. Dunn et al. reported a non-significant result (*P* = $`r format(wilcox_one_tailed1, digits= 3, scientific = FALSE)`$) for the PIC under the null simulation as well as for the empirical data, using a Wilcoxon one-tailed rank test to check if the contrasts of duplication events are higher than the contrasts of speciation events. Surprisingly, the PIC rejects the null hypothesis on the null simulations with a Wilcoxon two-tailed rank test (Fig 1A), with significant support for higher contrasts after speciation than duplication. This was robust to repeating the simulations with different random seed number (S1 Fig). This indicates that neither of the methods, PIC or pairwise, worked properly for these calibrated trees, since both the methods reject the null hypothesis when simulations are performed under the null. A two-tailed test is also significant with the empirical $\tau$ data using the sets of calibrated gene trees of Dunn et al., in the same direction as the null simulation results (median: PIC~speciation~ = `r round(median(speciation_empirical),4)`, PIC~duplication~ = `r round(median(duplication_empirical), 4)`, $`r format(wilcox_two_tailed_empirical, digits= 3, scientific = TRUE)`$).

@@

The purpose of applying a phylogenetic method is to generate independent statistical data points for traits, which are otherwise not independent because of their phylogenetic relatedness. Such statistical non-independence among species trait values can be measured by "phylogenetic signal" [@RN819; @RN768; @RN800; @RN820;@RN826]. Blomberg's K and Pagel's $\lambda$ are widely used methods to capture such phylogenetic signal [@RN819;@RN768]. Dunn et al. [@RN760] used Blomberg's K in their study. In case of Blomberg's K the value ranges from 0 to $\infty$, where a value of 0 indicates no phylogenetic signal, a value of 1 indicates trait evolution according to a Brownian model (BM), values between 0 and 1 indicate that distant relatives are more similar than expected under BM, and values larger than 1 indicate that close relatives are more similar than expected under BM [@RN819; @RN768; @RN800; @RN820;@RN826]. With a cutoff of K > 0.551, Dunn et al. [@RN760] obtained only `r subset.trees` out of `r length(gene_trees_calibrated)` calibrated trees with strong phylogenetic signal. Using a cut-off of *P* < 0.05 together with K > 0.551 leads to 1135 trees, which produce similar results (S3 Fig) to that set of `r subset.trees` trees. We continued analyses with the `r subset.trees` trees of Dunn et al. [@RN760] for consistency. In any case, it is notable that trait ($\tau$) values are independent of phylogeny for the majority of the gene trees. For these `r (length(gene_trees_calibrated)-subset.trees)` trees, implementation of PIC to produce statistically independent data points might not be necessary. Moreover, it can be misleading if the contrasts are not checked for adequate standardization as per BM after applying PIC (discussed later). The phylogenetic method still rejects the null hypothesis under null simulations for those `r subset.trees` trees (Fig 1B), showing that the problem is not simply due to low phylogenetic signal.

@@

```{r Deatiled_analyses_on_empirical_data, echo=F, message=F, warning=F}

@@

Dunn et al. [@RN760] used `r nrow(calibration_times_focal)` speciation time points to calibrate gene trees. The oldest speciation calibration was at `r max(calibration_times_focal$age)` My.  All other calibrated nodes were duplication nodes leading to obtain ~`r length(duplication.empirical)` unique duplication time points for the `r length(gene_trees_calibrated)` calibrated trees [@RN760]. Out of these, `r length(duplication.empirical[duplication.empirical > max(speciation.empirical)])` time points preceded the oldest speciation node age. Surprisingly, the calibrated node age of the oldest duplication event was found to be `r as.integer(max(duplication.empirical))` My, that is, 2600 times older than the Earth. This is indicative of the difficulty of estimating the age of ancient duplications by phylogenetic methods. Those `r length(duplication.empirical[duplication.empirical > max(speciation.empirical)])` high duplication node ages eventually led to much larger expected variances for gene duplication events (median expected variance for duplication events preceding the oldest speciation events: `r round(median(Exp_var.duplication.precede.speciation),0)`, median expected variance for speciation events: `r round(median(Exp_var.speciation),0)`). This explains why the mean paralog distance was ~`r round((paralog_distance_mean/ortholog_distance_mean),1)` times higher than that of the mean ortholog distance (S4A Fig). This data distribution also makes it appear as if no speciation happened before `r max(speciation.empirical)` My, a problem shared with the pairwise analysis of the same data [@RN762]. There are also branch length issues for duplication events younger than the oldest speciation events, including very short branches and negative branch lengths. All this led to obtain abnormally high duplication contrasts for the empirical data. To limit such problems, Dunn et al. [@RN760] removed node contrasts higher than 0.5 on the empirical data (this does not impact Fig 1, as the simulation data never has such high contrasts). Following this, there are still higher expected variances following gene duplication events (S4A and S4B Figs). In the null simulations only the $\tau$ values are simulated, while the branch lengths (hence the variances) are taken from the empirical data, and thus share its biases. This explains why contrasts are lower for duplications than for speciations under null simulations as well as with empirical data.

@@

  ## All calibrated trees with permuted actual Tau of the tips 
  set.seed(23456)
  Tau_randomized_tree <- mclapply(gene_trees_pic,shuffling_tau,mc.cores = cores)

@@

  ## All calibrated trees with permuted actual internal node events
  set.seed(23456)
  Event_randomized_pic_tree <- mclapply(gene_trees_pic, shuffling_event, mc.cores = cores)

@@

  ## Histogram to show distribution of proportion of events for all calibrated trees
  event_histogram <- ggplot2_histogram_mod(data=plot_data, xName='proportion_event',
    groupName='Event',legendPosition="right",
    alpha=1, addDensity=TRUE,
    addMedianLine=TRUE, medianLineColor=c("#F8766D","#00BFC4"), medianLineSize=1) +
    theme_classic() +
    ggtitle(expression(bold("D. Event proportion plot")))+
    theme(plot.title = element_text(face="bold"))+
    xlab("Proportion of events") + 
    ylab("Density") +
    theme(legend.title=element_blank(),legend.position=c(0.8,0.7),legend.text=element_text(size=8)) +
    theme(axis.text=element_text(size=10,face="bold")) +
    theme(plot.title = element_text (face="bold", 
                              colour="black", lineheight=1.0),
          axis.title.x=element_text( face="bold",
                              colour="black", hjust=0.5),
          axis.title.y=element_text(face="bold",
                              colour="black", vjust=0.5, angle=90))

@@

We used randomization tests to assess whether the results of different analyses of the empirical dataset are reliable and unbiased. Our expectation is that results from the empirical result should differ from the randomized ones. In a first randomization test, we randomized the trait values (i.e. randomly permuting the $\tau$ values) across the tips of each tree without altering the node events of the trees. We then computed the contrasts for the speciation and duplication events of the trees when there is no relation between trait values and phylogenetic relationships. When we compared the node contrasts of speciation and duplication events of these randomized trees (Fig 2A), we found the same pattern as reported for the empirical gene trees by Dunn et al. [@RN760]. This shows that the PIC applied to these data is not measuring phenotypic evolution, and thus not meeting the aim for applying the method. It confirms that results are driven by their large differences in branch lengths (i.e. in expected variances) (Fig 2B), as on simulated null data. Any effect of trait divergence rates of speciation and duplication events is masked by this branch length difference. This violates the basic assumption of applicability of the PIC method to Brownian trait evolution. To remove the problem of difference in expected variances of the two events, we performed a second randomization test: we kept the original $\tau$ value for tips but randomly shuffled the events (duplication or speciation) of internal nodes of the empirical gene trees to maintain the original proportions of speciation and duplication events. The resulting trend (Fig 2C) still resembled the empirical gene trees data. This appears due to the fact that the majority of the tree events are speciation events (Fig 2D) with speciation node ages $\leq$ 296 My. Most of the trees with many duplication events on the other hand have ancient duplication events (duplication node age > 296 My) for which the evolutionary rates of duplication are often masked by the effect of longer branch lengths. Hence, opposite to our expectation, the calibrated trees with no or few duplications have higher overall nodes contrast (apparent fast evolution) than trees with many duplications (apparent slow evolution). This might be due to greater difficulty in detecting paralogs for fast evolving genes. Therefore, reshuffling of the events may not change the observed pattern of higher speciation contrasts than duplication contrasts.

@@

```{r Fig2, dpi=400,fig.width=8, fig.height=8, fig.cap="**Fig 2**: Analyses on calibrated empirical gene trees of Dunn et al. [@RN760]. *P* values are from  Wilcoxon two-tailed tests. (A) Randomly shuffling the $\\tau$ values of the tips for 8520 gene trees does not alter the trend of empirical result [@RN760] showing an opposite trend to the ortholog conjecture. (B) The figure indicates the expected variance is much higher for duplication than speciation events irrespective of the number of tips considered for the study. (C) Using the original $\\tau$ data, if we permute the events (Speciation or Duplication or NA) of the nodes, the trend of result still resembles the empirical result obtained by the recent study [@RN760]. (D) The proportions of speciation events are much higher than duplication events for all time-calibrated trees. The dotted line represents the median proportion of both events. This plot shows a high frequency of trees have no/few duplication events."}

  ## Plotting the randomization data result for all calibrated trees 
  ## 1st plot
  jitterplot_randomized_tau <- 
    ggplot(nodes_contrast_random,aes(x=Event, y=pic_abs_random, fill=Event)) + 
    geom_jitter(aes(colour = Event),position=position_jitter(0.02), alpha=0.02) +
    geom_text(data = median_nodes_contrast_random, aes(label = pic.round),fontface = 2,size = 4, hjust=0.5,vjust =-2.5)+
    geom_crossbar(data=median_nodes_contrast_random, aes(ymin = pic_abs_random, ymax = pic_abs_random),size=0.2,col= "black", width = .2)+
    xlab( NULL ) +
    ylab(expression(bold("Phylogenetic Independent Contrasts (PICs)"))) +
    ylim(0, 0.4) +
    theme_classic()+
    theme(legend.title=element_blank(),legend.position=c(1.9,1.9)) +
    theme(axis.text=element_text(size=10,face="bold")) +
    theme(legend.text=element_text(size=10,face="bold")) +
    annotate("text", x = 1.5, y = 0.3, label= wilcox_2_tailed.r, fontface = 4)+
    ggtitle(expression(bold(paste("A. Plot using randomized trait, ", tau))))+
    theme(plot.title = element_text(face="bold"))

@@

  jitterplot_randomized_event <- 
    ggplot(nodes_contrast_random_events,aes(event_new, y=pic, fill=event_new)) + 
    geom_jitter(aes(colour = event_new),position=position_jitter(0.02), alpha=0.02) +
    geom_text(data = median_random_event, aes(label = pic.round),fontface = 2,size = 4, hjust=0.5,vjust = -2.5)+
    geom_crossbar(data=median_random_event, aes(ymin = pic, ymax = pic),
                  size=0.2, col= "black",width = .2)+
    xlab( NULL ) +
    ylab( NULL) +
    ylim(0, 0.4) +
    theme_classic()+
    theme(legend.title=element_blank(),legend.position=c(1.8,1.8)) +
    theme(axis.text=element_text(size=10,face="bold")) +
    theme(legend.text=element_text(size=10,face="bold")) +
    annotate("text", x = 1.5, y = 0.3, label= wilcox_2_tailed_revent, fontface = 4)+
    ggtitle(expression(bold(paste("C. Plot using randomized events"))))+
    theme(plot.title = element_text(face="bold"))

@@

__The ortholog conjecture when $\tau$ evolution follows a non-Brownian model __

@@

```{r Analyses_on_standardized_lambda_transformed_3879_trees, eval=F, echo=F, message=F, warning=F}

  ##summarizing
   summary_brownian_3879 <- bind_rows(lapply(standarized_new_Brownian,summary_function))
   Contrast_brownian_3879_trees <- summary_brownian_3879[which(!is.na(summary_brownian_3879$pic_Tau)),]
